---
title: "Untitled"
author: "Jungin Choi"
date: "2024-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, message = FALSE, warning=FALSE}
library(LogConcDEAD)

SPMix <- function(z, init_p0 = 0.5, tol = 5.0e-5, np.left = NULL,
                  alternative = "greater", min_iter = 3, max_iter = 50, 
                  thre_z = 0.5, Uthre_gam = 0.99, Lthre_gam = 0.01 )
{
  # *****************DEFINITION OF INTERNAL FUNCTIONS ******************
  
  NormMix <- function(z, p0, tol = 5e-3, max_iter = 5)
  {
    require(mvtnorm)
    
    k <- 0; converged <- 0
    z <- as.matrix(z)
    n <- nrow(z)
    d <- ncol(z)
    q <- qnorm(p0)
    z.scaled <- scale(z)
    ind0 <- ind1 <- rep(1, n)
    for (j in 1:d) {
      ind0 <- ind0*(z.scaled[,j] <= q)
      ind1 <- ind1*(z.scaled[,j] > -q)
    }
    
    z0 <- as.matrix(z[ind0 == 1,])
    z1 <- as.matrix(z[ind1 == 1,])
    
    mu0 <- colMeans(z0)
    sig0 <- cov(z0)
    f0 <- dmvnorm(z, mu0, sig0)
    
    mu1 <- colMeans(z1)
    sig1 <- cov(z1)
    f1 <- dmvnorm(z, mu1, sig1)
    
    f <- p0*f0 + (1 - p0)*f1
    ell <- mean(log(f))
    
    while ((k < 3) | ((k < max_iter) & (!converged))) {
      k <- k + 1
      
      ## E-step
      gam <- p0 * f0 / (p0 * f0 + (1 - p0) * f1)
      
      ## M-step
      new_p0 <- mean(gam)
      new_mu0 <- as.vector(t(z) %*% gam) / sum(gam)
      dev0 <- (z - new_mu0) * sqrt(gam)
      new_sig0 <- t(dev0) %*% dev0 / sum(gam)
      f0 <- dmvnorm(z, new_mu0, new_sig0)
      
      new_mu1 <- as.vector(t(z) %*% (1 - gam)) / sum(1 - gam)
      dev1 <- (z - new_mu1) * sqrt(1 - gam)
      new_sig1 <- t(dev1) %*% dev1 / sum(1 - gam)
      f1 <- dmvnorm(z, new_mu1, new_sig1)
      
      f <- p0*f0 + (1 - p0)*f1
      new_ell <- mean(log(f))
      
      ## Update
      diff <- abs(new_ell - ell)
      converged <- (diff <= tol)
      p0 <- new_p0
      mu0 <- new_mu0
      sig0 <- new_sig0
      mu1 <- new_mu1
      sig1 <- new_sig1
      ell <- new_ell
    }
    
    return(list(p0 = p0, mu0 = mu0, sig0 = sig0, mu1 = mu1, sig1 = sig1))
  }
  
  NE <- function(x, X)
  {
    n <- nrow(X)
    p <- ncol(X)
    xx <- matrix(x, nrow = n, ncol = p, byrow = TRUE)
    ne.ind <- apply(1*(X >= xx), 1, prod)

    return((1:n)[ne.ind == 1])
  }

  MonotoneFDR <- function(z, fdr)
  {
    n <- nrow(z)
    MFDR <- numeric(n)
    for (i in 1:n) {
      MFDR[i] <- max(fdr[NE(z[i,], z)])
    }

    return(MFDR)
  }
  
  mult.ecdf <- function(x)
  {
    x <- as.matrix(x)
    n <- nrow(x)
    d <- ncol(x)
    Fn <- rep(0, n)
    for ( i in 1:n ) {
      tmp <- matrix(x[i,], n, d, byrow = TRUE)
      Fn[i] <- mean(rowSums(x <= tmp) == d)
    }
    return(Fn)
  }

  # ******************* MAIN FUNCTION *******************************
  
  z <- as.matrix(z)
  n <- nrow(z)
  d <- ncol(z)
  ell <- rep(NA, max_iter)
  
  if(!is.null(np.left)) {
    for(j in 1:d) {
      if(np.left[j] == "Yes") z[,j] <- -z[,j]
    }
  }
  
  ## Initial step: to fit normal mixture
  if (d == 1) {
    z = z[,1]
    if (alternative == "greater" | alternative == "g") {
      q0 <- quantile(z, probs = .9)
      p0 <- mean(z <= q0)
      mu0 <- mean(z[z <= q0])
      sig0 <- sd(z[z <= q0])
      f0 <- mclust::dmvnorm(z, mu0, sig0)
      mu1 <- mean(z[z > q0])
      sig1 <- sd(z[z > q0])
      f1 <- dnorm(z, mu1, sig1)
    } else {
      q0 <- quantile(z, probs = .7)
      p0 <- mean(z >= q0)
      mu0 <- mean(z[z >= q0])
      sig0 <- sd(z[z >= q0])
      f0 <- mclust::dmvnorm(z, mu0, sig0)
      mu1 <- mean(z[z < q0])
      sig1 <- sd(z[z < q0])
      f1 <- dnorm(z, mu1, sig1)
    }
  } else {
    ## Initial step: to fit normal mixture
    Params <- NormMix(z, p0 = init_p0)
    p0 <- Params$p0
    mu0 <- Params$mu0
    sig0 <- Params$sig0
    f0 <- dmvnorm(z, mu0, sig0)
    f1 <- dmvnorm(z, Params$mu1, Params$sig1)
    f <- p0*f0 + (1 - p0)*f1
    ell[1] <- mean(log(f), na.rm = TRUE)
  }
  gam <- f <- rep(0, n)

  if (d == 1) {
    z <- as.numeric(z)
    ## EM-step
    k <- 0; converged <- 0
    while ( (k < min_iter) | ((k < max_iter) & (!converged)) ) {
      k <- k + 1

      ## E-step
      new_f <- (p0 * f0 + (1 - p0) * f1)
      new_gam <- p0 * f0 / new_f

      ## M-step

      w_gam <- new_gam/sum(new_gam, na.rm = TRUE)
      new_mu0 <- sum(w_gam*z, na.rm = TRUE)
      new_sig0 <- sqrt(sum(w_gam*(z-new_mu0)^2, na.rm = TRUE))
      new_p0 <- mean(new_gam, na.rm = TRUE)
      new_f0 <- dnorm(z, new_mu0, new_sig0)
      new_f1 <- rep(0, n)
      which_z <- (new_gam <= thre_z)
      weight <- 1 - new_gam[which_z]
      weight <- weight/sum(weight)
      new_f1[which_z] <- exp(LogConcDEAD::mlelcd(z[which_z], w = weight)$logMLE)

      ## Update
      which_gam <- (new_gam <= Uthre_gam) * (new_gam >= Lthre_gam)
      diff <- max(abs(gam - new_gam)[which_gam])
      converged <- (diff <= tol)
      cat("   EM iteration:", k, ", Change in fdr fit = ", round(diff, 5), "\n")
      p0 <- new_p0
      mu0 <- new_mu0
      sig0 <- new_sig0
      f1 <- new_f1
      f0 <- new_f0
      f <- p0 * f0 + (1 - p0) * f1
      gam <- new_gam

    }
  } else {
    ## EM-step
    k <- 1; converged <- 0
    while ( (k < 5) | ((k < max_iter) & (!converged)) ) {
      k <- k + 1
      
      ## E-step
      gam <- p0 * f0 / f
      gam <- MonotoneFDR(z, gam)
      
      ## M-step
      sum_gam <- sum(gam)
      mu0 <- as.vector(t(z) %*% gam) / sum_gam
      dev <- t(t(z) - mu0) * sqrt(gam)
      sig0 <- t(dev) %*% dev / sum_gam
      p0 <- mean(gam)
      f0 <- dmvnorm(z, mu0, sig0)
      f1 <- rep(0, n)
      which_z <- (gam <= thre_z)
      weight <- 1 - gam[which_z]
      weight <- weight/sum(weight)
      #lcd <- fmlogcondens::fmlcd(X = z[which_z,], w = weight[which_z] / sum(weight[which_z]))
      lcd <- LogConcDEAD::mlelcd(z[which_z,], w = weight)
      f1[which_z] <- exp(lcd$logMLE)
      f <- p0*f0 + (1 - p0)*f1
      ell[k] <- mean(log(f), na.rm = TRUE)
      
      ## Update
      diff <- abs(ell[k] - ell[k - 1])
      converged <- (diff <= tol)
      cat(".")
      if (converged) {
        cat("Converged!\n")
        break
      }
    }
    if(!converged) cat("Warning: Not converged!\n")
  }

  # return results

    if (d == 1){
      res <- list(z = z*raw_sd + raw_mean, p0 = p0, mu0 = mu0*raw_sd + raw_mean, 
                  sig0 = sig0*raw_sd, f = f/raw_sd, f1 = f1/raw_sd, 
                  localFDR = gam, iter = k, dim = d, alternative = alternative)
    } else {
      # return results
      res <- list(z = z, p0 = p0, mu0 = mu0, sig0 = sig0, f1 = f1, f = f,
                  iter = k, log.likelihood = ell, lcd = lcd, dim = d,
                  posterior = cbind(gam, 1 - gam),
                  converged = converged)
    }
  
  return(res)
}





plotSPMix <- function(x, thre_localFDR = 0.2, testing = FALSE,
                      xlab = "x", ylab = "y", zlab = "z", coord_legend = c(8, -5, 0.2))
{
  z <- x$z
  p0 <- x$p0
  mu0 <- x$mu0
  sig0 <- x$sig0
  f <- x$f
  f1 <- x$f1
  d <- x$dim
  
  if(!is.null(np.left)) {
    for(j in 1:d) {
      if(np.left[j] == "Yes") x$mu0[j] <- -x$mu0[j]
    }
  }
  

  if (d == 1){
    z = as.numeric(z)
    if (alternative == "greater" | alternative == "g"){
      thre <- min(z[which_z])
    } else{
      thre <- max(z[which_z])
    }

    legend_testing <- factor(which_z, levels = c(FALSE, TRUE), labels = c("Nonsignificant", "Significant"))
    legend_density <- factor((localFDR <= 0.5), levels = c(FALSE, TRUE), labels = c("Normal", "Nonparametric"))

    sub_testing=substitute(
      paste("Hypothesis Testing: ", p[0], " = ", p0, ", ",
            mu[0], " = ", mu0, ", ",
            sigma[0], " = ", sigma0, ", ",
            "threshold = ", threshold,
            sep = ""),
      list(p0 = round(p0, 2),
           mu0 = round(mu0, digits = 2),
           sigma0 = round(sig0, digits = 2),
           threshold = round(thre, digits = 2)))

    sub_density=substitute(
      paste("Density Estimation: ", p[0], " = ", p0, ", ",
            mu[0], " = ", mu0, ", ",
            sigma[0], " = ", sigma0,
            sep = ""),
      list(p0 = round(p0, 2),
           mu0 = round(mu0, digits = 2),
           sigma0 = round(sig0, digits = 2)))

    df = data.frame(z=z)
    zs <- sort(z)
    if (testing) {
      ggplot(df,aes(x=z)) +
        geom_histogram(aes(y = ..density..,),colour = 1, fill = "white", bins=100) +
        geom_line(aes(sort(z), (f[order(z)])),color = "gray25", lwd=1.1) +
        geom_line(aes(sort(z), p0*dnorm(zs, mean = mu0, sd = sig0)),color = "#0072B2",lwd=0.7) +
        geom_line(aes(sort(z), ((1-p0)*f1[order(z)])),color = "#D55E00",lwd=0.7) +
        geom_vline(aes(xintercept=thre), color="#E69F00",linetype="dashed") +
        geom_point(mapping = aes(x = thre, y = 0.01),size = 2,color="#E69F00",shape=25, fill="#E69F00") +
        labs(x=xlab, y = "density") +
        ggtitle(sub_testing) +
        theme(plot.title = element_text(margin = margin(b = -10))) +
        geom_rug(aes(z,color = legend_testing)) +
        scale_color_manual(values = c("#999999", "#E69F00"), name="") +
        theme_classic()
    } else {
      ggplot(df,aes(x=z)) +
        geom_histogram(aes(y = ..density..),colour = 1, fill = "white",bins=100) +
        geom_line(aes(sort(z), (f[order(z)])),color = "gray25", lwd=1.1) +
        geom_line(aes(sort(z), p0*dnorm(zs, mean = mu0, sd = sig0)),color = "#0072B2",lwd=0.7) +
        geom_line(aes(sort(z), ((1-p0)*f1[order(z)])),color = "#D55E00",lwd=0.7) +
        labs(x=xlab, y = "density") +
        ggtitle(sub_density) +
        theme(plot.title = element_text(margin = margin(b = -10))) +
        geom_rug(aes(z,color = legend_density)) +
        scale_color_manual(values = c("#0072B2", "#D55E00"), name="") +
        theme_classic()
    }

  } else if (d == 2) {
    sub_testing=substitute(
      paste("Hypothesis Testing: ", p[0], " = ", p0, ", ",
            mu[0], " = (", mu01,",",mu02, "), ",
            "localFDR threshold = ", threshold,
            sep = " "),
      list(p0 = round(p0, 2),
           mu01 = round(mu0[1], digits = 2),
           mu02 = round(mu0[2], digits = 2),
           threshold = round(thre_localFDR, digits = 2)))

    sub_density=substitute(
      paste("Density Estimation: ", p[0], " = ", p0, ", ",
            mu[0], " = (", mu01,",",mu02, "), ",
            sep = ""),
      list(p0 = round(p0, 2),
           mu01 = round(mu0[1], digits = 2),
           mu02 = round(mu0[2], digits = 2)))

    sub_3d <- if (testing) {sub_testing} else {sub_density}
    
    if (!testing) {
      # Contour Plot
      ngrid <- 50
      
      x1 <- seq(from = min(z[,1]), to = max(z[,1]), length = ngrid)
      x2 <- seq(from = min(z[,2]), to = max(z[,2]), length = ngrid)
      
      comp0 <- x$p0 * dmvnorm(as.matrix(expand.grid(x1, x2)), x$mu0, x$sig0)
      comp1 <- (1 - x$p0) * dlcd(as.matrix(expand.grid(x1, x2)), x$lcd, uselog = FALSE)
      comp0 <- matrix(comp0, ngrid, ngrid)
      comp1 <- matrix(comp1, ngrid, ngrid)
      den <- comp0 + comp1
      
      layout(matrix(1))
      
      image(x1, x2, den, cex.axis = 0.7, xlab = xlab, ylab = ylab, 
            main = sub_3d, col = hcl.colors(50))
      points(x$z, pch = 20, col = "gray")
      contour(x1, x2, comp0, add = TRUE, col = "white")
      contour(x1, x2, comp1, add = TRUE, col = "white")
    } else{
      z <- x$z
      x$local.fdr <- x$posterior[,1]
      discovered <- (x$local.fdr <= thre_localFDR)  + 1
      library(scales)
      
      ngrid <- 50
      x1 <- seq(from = min(z[,1]), to = max(z[,1]), length = ngrid) 
      x2 <- seq(from = min(z[,2]), to = max(z[,2]), length = ngrid)
      par(mfrow = c(1, 2))
      comp0 <- x$p0*dmvnorm(as.matrix(expand.grid(x1, x2)), x$mu0, x$sig0)
      comp0 <- matrix(comp0, ngrid, ngrid)
      comp1 <- (1 - x$p0)*dlcd(as.matrix(expand.grid(x1, x2)), x$lcd, uselog = FALSE)
      comp1 <- matrix(comp1, ngrid, ngrid)
      den <- comp0 + comp1
      image(x1, x2, den, cex.axis = 0.7, 
            xlab = "", ylab = "", col = hcl.colors(50))
      cols <- c("#999999", "#E69F00")[discovered]
      points(z, pch = 20, col = alpha(cols, 0.4))
      contour(x1, x2, comp0, add = TRUE)
      contour(x1, x2, comp1, add = TRUE)
      
      plot(x$local.fdr, xlab = "index", ylab = "local fdr", 
           pch = 20, col = cols)
      abline(h = thre_localFDR, col = 2, lty = 2)
    }
  
  } else if (d == 3) {
    colors <- c("#999999", "#E69F00")
    colors <- colors[as.numeric(which_z)+1]
    scatterplot<- scatterplot3d(z[,1],z[,2],z[,3], pch = 16, color=colors,
                                xlab = xlab, ylab = ylab, zlab = zlab)
    legend_testing <- factor(which_z, levels = c(FALSE, TRUE), labels = c("Nonsignificant", "Significant"))
    legend_density <- factor((localFDR <= 0.5), levels = c(FALSE, TRUE), labels = c("Normal", "Nonparametric"))
    legend_3d <- if (testing) {legend_testing} else {legend_density}
    legend(scatterplot$xyz.convert(coord_legend[1], coord_legend[2], coord_legend[3]),
           legend = levels(legend_3d), col = c("#999999", "#E69F00"), pch = 16)
  }
}

```

# Simulations

In this section, we will use SPMix() to estimate mixture density and compute local-FDR for generated data. Also, plotSPMix() can visualize the fitted results for both univariate and 2-dimensional data.

## required libraries
```{r warning=FALSE, message=FALSE}
library(copula)
library(mvtnorm)
library(ggplot2)
library(scatterplot3d)
```


## 2-dimensional data

### Normal + Gaussian Copula

```{r}

n <- 3000
p0 <- 0.8
n0 <- rbinom(1, n, p0)
n1 <- n - n0
sig0 <- matrix(c(1, 0.25, 0.25, 1), ncol = 2, byrow = T)
V <- rCopula(n1, ellipCopula(family = "normal", param = 0.5, dim = 2))
z0 <- rmvnorm(n0, sigma = sig0)
z1 <- qnorm(V, mean = 3.5, sd = .5)
z_NN2d <- data.frame(rbind(z0, z1))

ggplot(z_NN2d, aes(x = z_NN2d[,1], y = z_NN2d[,2])) +
  geom_point(aes(color = factor(c(rep(1,n0), rep(2,n1)), labels = c("Normal(Null)", "Gaussian Copula(Alternative)")))) +
  scale_color_manual(values = c("#999999", "#E69F00"), name="")  +
  labs(title = "Random Generated Mixture density data", x = "x", y = "y") +
  theme_classic()
```


```{r}
NN2d <- SPMix(z_NN2d)

plotSPMix(NN2d)
plotSPMix(NN2d, testing = TRUE)
```

__plotSPMix()__ also provides visualization of 2-dimensional data.
You can set name for x-axis and y-axis using __xlab__, __ylab__. 

### Normal + Frank Copula

Let's consider nonnull distribution as Frank copula with marginal gamma distribution.

```{r eval = FALSE}
z1_gamma <- qgamma(V, shape = 12, rate = 4)
z_NG <- data.frame(rbind(z0, z1_gamma))

z1_gamma <- qgamma(V, shape = 10, rate = 4)
z_NG2d <- data.frame(rbind(z0, z1_gamma))

ggplot(z_NG, aes(x = z_NG[,1], y = z_NG[,2])) +
  geom_point(aes(color = factor(c(rep(1,n0), rep(2,n1)), labels = c("Normal", "Gamma")))) +
  scale_color_manual(values = c("#999999", "#E69F00"), name="")  +
  labs(title = "Random Generated Normal/Gamma Mixture", x="x", y = "y") +
  theme_classic()
```

```{r eval = FALSE}
NG <- SPMix(z_NG)

plotSPMix(z_NG, NG$p0, NG$mu0, NG$sig0, NG$f, NG$f1, NG$localFDR, 
          xlab = "x", ylab = "y")

plotSPMix(NG2d)
```
